---
title: "rHFA Guide"
author: "Erik Killian"
date: "2022-09-12"
output: html_document

---

```{r setup, include=FALSE, eval=FALSE,echo=FALSE}
# knitr::opts_chunk$set
# install.packages(c("devtools", "roxygen2", "car", "lme4"))
# install.packages(c("filter_siteyears", "get_home_site", "get_libraries", "get_SS", "id_home", "permute_hfa", "temporal_hfa"))
library(devtools)
library(roxygen2)
library(car)
library(lme4)
library(rHFA)
```

# Introduction

## Home Field Advantage in Plant Breeding

Background of the HFA stat and why it's important.

### How Home Field Advantage is Calculated

HFA equation and a description. Could add some example data/code?

# Guide

## Step 1: Load dataset



## Step 2: ID home site

```{r eval=FALSE}
id_home(df = data_yield, year = "YEAR", geno = "HYBRID", pheno = "yield")
```


```{r id_home}
.scale_pheno = function(x, pheno) {
  # for applying
  colname = paste0('rel_', pheno)
  x[, colname] = scale(x[, pheno])
  return(x)
}

#' @title .id_top_pheno
#' 
#' @description id the level of `x[, site]` where the mean relative value of 
#' `x[, pheno]` is highest. By default, uses shrinkage estimates of mean 
#' relative `x[, pheno]` (via `lme4`) if any site occurs in at least two 
#' years. Otherwise (or optionally) uses means.
#' 
#' @param x data.frame or matrix for a single unit, ex a genotype. Coerced to data.frame.
#' @param site character of grouping column, ex field site locations
#' @param pheno character of response, ex yield, biomass
#' @param blup logical - use shrinkage estimates?
#' @param verbose logical
#' 
#' @returns x, with an additional logical column `"Home"`, which indicates 
#' whether the value of `x[, site]` is the home site.

.id_top_pheno = function(x, site, pheno, blup=TRUE, verbose=TRUE) {
  require(magrittr)
  
  x %<>% as.data.frame(stringsAsFactors=FALSE)
  
  x[, site] %<>%
    factor %>%
    droplevels
  
  # find max years at any one site. 
  if (blup) {
    site_years = x[, site] %>%
      tapply(., ., length) %>%
      max
    if (site_years < 3) { #(site_years < 2 & blup) {#
      if (verbose) message('Cannot blup home field. Using (unshrunk) mean values')
      blup = FALSE
    }
  }
  
  # calculate mean pheno within site.
  if (blup) {
    require(lme4)
    
    if (verbose) message('Using lme4 to identify the home site')
    site_eff = paste(pheno, '~ 0 + (1|', site, ')') %>% 
      formula %>% 
      lmer(data=x,
           control=lmerControl(calc.derivs = FALSE)) %>% # tip from lme4 pheno vignette
      coef %>% 
      extract2(site)
    site_eff %<>% 
      .[,1] %>% 
      set_names(rownames(site_eff))
    
  } else {
    site_eff = tapply(x[, pheno], x[, site], mean, simplify=TRUE)  # faster than aggregate
  }
  
  # identify site with max mean
  is_max = site_eff %>% 
    which.max
  max_site = names(site_eff)[is_max]
  
  x$is_home = x[, site] == max_site
  
  return(x)
}

#' @title id_home
#' 
#' @description Identify the home site based on highest relative phenotype value. 
#' 
#' @param df data.frame of performance data by site, year, and variety, for example. 
#' @param site character, column name indicating spatial units
#' @param year character, column name indicating temporal units
#' @param geno character, column name indicating genetic units
#' @param pheno character, column name indicating the phenotype (ex. yield)
#' @param blup logical (`TRUE`). Use shrinkage estimates of mean relative performance within each site across years?
#' @param verbose logical (`TRUE`).
#' 
#' @details The home site for a given variety is defined as the location where 
#' variety performs best across years, relative to other varieties. It's calculated
#' by:
#' 
#' 1. Calculate relative phenotype within site-year by scaling (mean = 0) and 
#' scaling (st. dev. = 1) phneotype across varieties within site-years
#' 2. Find the expected relative phenotype for each variety within
#' within sites across years. This is either the raw mean or the shrunk/BLUP mean (via
#' a random intercept model using `lme4`). 
#' 3. For each variety, identify the site with the highest expected relative phenotype value. 
#' 
#' @returns a data.frame with two new columns:
#' 
#' - `'rel_<pheno>'`: numeric, the relative phenotype value of each variety within site-year
#' - `is_home`: logical of whether a site is a variety's home site.

id_home = function(df, site, year, geno, pheno, blup=TRUE, verbose=FALSE) {
  require(magrittr)
  require(parallel)
  
  rel_colname = paste0('rel_', pheno)
  
  # make site-year vector
  site_year = paste(df[, site], df[, year], sep='_')
  # center/scale performance within site-year
  df %<>% 
    split(site_year) %>% 
    lapply(.scale_pheno, pheno) %>% 
    do.call(rbind, .)
  
  # find highest relative yield for each genotype
  df %<>%
    split(df[geno]) %>%
    lapply(.id_top_pheno, site, rel_colname, blup, verbose) %>% 
    do.call(rbind, .)
  
  return(df)
}

```  

```{r get_home_site}
#' @title get_home_site
#' 
#' @description extract home location information for each genotype
#' 
#' @param data data.frame from call to `id_home()`
#' @param geno column in `data` with genotype
#' @param site column in `data` with location environment
#' 
#' @return a `data.frame` with columns `geno`, `site`, and `home_years` - the number
#' of years a genotype appeared at a home site.

get_home_site = function(data, geno, site) {
  is_home = data[, 'is_home']
  out = data[is_home, c(geno, site)]
  counts = out[, geno] %>% 
    tapply(., ., length)
  
  out %<>% unique
  out[, 'home_years'] = counts[out[, geno]]
  return(out)
}


#' @title measure_home_distance
#' 
#' @description calculate the distance between home sites for among genotypes
#' 
#' @param data data.frame from a call to `id_home()`
#' @param locations a SpatialPointsDataFrame (recommended) or a data.frame with columns `site`, `lat`, and `long`.
#' @param geno genotype column in `data` and in `locations`
#' @param site spatial environment column in `data`
#' @param lat latitude/y coordinate column in `locations`
#' @param long longitude/x coordinate column in `locations`
#' @param great_circle are coordinates on an ellipse? 
#' 
#' @return a distance matrix. row/column names are genotypes. Values are distances
#' among between home sites of the genotype pairs.
#' 
#' @note If lat/long are in degrees, then set `great_circle` to TRUE. If lat/long
#' are in linear units, set false. Can also pass a SpatialPointsDataFrame, which will
#' override this the `great_circle` setting and coordinates will be derived from the object. 
#' 
#' @seealso package `"sp"`

measure_home_distance = function(data, locations, geno, site, lat=NA, long=NA, great_circle=TRUE) {
  require(sp)
  require(magrittr)
  
  if (grepl("SpatialPoints", class(locations))) {
    distances = spDists(locations)
  } else {
    distances = locations[, c(long, lat)] %>% 
      as.matrix %>% 
      spDists(longlat=great_circle)
  }
  
  colnames(distances) = locations[[site]]
  rownames(distances) = locations[[site]]
  
  data %<>% 
    get_home_site(geno, site)
  
  vars = data[, geno] %>% 
    as.character
  homes = data[, site] %>%
    as.character %>% 
    set_names(vars)
  nvar = length(vars)
  
  out = matrix(nrow = nvar,
               ncol = nvar) %>% 
    set_rownames(vars) %>% 
    set_colnames(vars)
  
  for (i in vars) {
    home_i = homes[i]
    for (j in vars) {
      home_j = homes[j]
      out[i, j] = distances[home_i, home_j]
    }
  }
  
  return(out)
}

```

```{r get_libraries}
#' @title try to load a library, and download if needed
#' 
#' @description mostly helps resolve packrat bugs. Also installs in parallel.
#' 
#' @param lib_string charracter (vector) with package names
#' @param cores integer or logical - install packages in parallel? TRUE uses
#' `parallel::detectCores()`
#' 
#' @returns nothing, but loads packages into the user environment.

get_libraries = function(lib_string, in_parallel=TRUE) {
  
  if (is.logical(in_parallel)) {
    tt = in_parallel & ('parallel' %in% rownames(installed.packages()))
    if (tt) {
      threads = parallel::detectCores() 
    } else {
      threads = 1
    }
    
  } else {
    threads = in_parallel
  }
  
  if (!require(lib_string, character.only=TRUE)) {
    msg = paste('Trying to install', lib_string)
    message(msg)
    
    install.packages(lib_string, Ncpus=threads)
    if (!require(lib_string, character.only=TRUE)) {
      msg = paste("Could not install", lib_string)
      stop(msg)
    }
  }
}

```

```{r filter_siteyears}
#' @title filter_siteyears
#' 
#' @description Subset locations by a minimum number of occurrances
#' 
#' @param x data.frame
#' @param site column containing trial location
#' @param year column containing year
#' @param min_times numeric, minimum number of years a site must be present. Default 3. 
#' 
#' @return a data.frame, a subset of x. 

filter_siteyears = function(x, site, year, min_times=3) {
  require(magrittr)
  
  common_sites = c(site, year) %>% 
    x[, .] %>% 
    unique %>%
    .[[site]]
  
  common_sites %<>%
    tapply(., ., length) %>%  # count occurrences
    .[!is.na(.)]
  
  common_sites %<>%  # filter by min_times
    .[.>=min_times]
    

  test = x[, site] %in% names(common_sites)
  out = x[test, ]
  
  return(out)
}

```

```{r get_SS}
get_ss = function(model) {
  require(car) 
  
  a = Anova(model)
  out = data.frame(
    PREDICTOR = rownames(a),
    SUMSQ = a$`Sum Sq`,
    pVAR = round(a$`Sum Sq` / sum(a$`Sum Sq`) *100, 2),
    F_val = round(a$`F value`, 4),
    p_val = signif(a$`Pr(>F)`, 3)
  )
  return(out)
}


```

```{r permute_hfa}
#' @title .quick_resid <internal>
#' 
#' @description calculate residuals of a linear model as efficiently as possible (using base R).
#' Residuals can be used in subsequent regressions as partial effects.
#' 
#' @param ff formula
#' @param data data.frame containing formula
#' 
#' @return A vector of residuals, length = nrow(data)

.quick_resid = function(ff, data) {
  require(Matrix)

    X = model.matrix(ff, data) 
  X = qr(X)
  Y = model.frame(ff, data)[, 1]
  
  out = qr.resid(X, Y)

  return(out)
}


#' @title .generate_sets <internal>
#' 
#' @description generate permutation sets structured by site-year
#' 
#' @param x data.frame
#' @param site column identifying spatial environment
#' @param year column identifying temporal environment
#' @param times number of sets to produce. 
#' 
#' @return a data.frame of permutation sets - row numbers used to reorder x. The first column is the original data.

.generate_sets = function(x, site, year, times, seed=NULL) {
  require(magrittr)
  require(permute)
  
  control = paste(x[, site], x[, year], sep='_') %>% 
    as.factor
  N = nrow(x)
  
  set.seed(seed)
  ss = shuffleSet(N, times, control=how(blocks=control)) %>% # permutations
    rbind(seq_len(N),  # include original data
          .) %>% 
    t %>% 
    data.frame
  
  return(ss)
}

#' @title .calculate_hfa <internal>
#' 
#' @description calculate the home field advantage based on the formula, ff. Uses efficient
#' packages for fastest permutation results.
#' 
#' @param x data.frame
#' @param ff formula to use. part_pheno ~ geno */+ is_home
#' 
.calculate_hfa = function(x, ff, part_pheno, geno, year, LAPACK=TRUE) {
  require(Matrix)
  
  X = droplevels(x) %>%   # make sure all factors in ff are present for LAPACK
    model.matrix(ff, data=.) %>% 
    qr(LAPACK=LAPACK)
  
  Y = x[, part_pheno, drop=FALSE]
  
  home_coef = qr.coef(X, Y) %>% 
    as.matrix
  
  is_home = grepl('is_homeTRUE', rownames(home_coef))
  home_coef %<>% .[is_home, , drop=FALSE]
  
  rownames(home_coef) %<>% 
    gsub(':is_homeTRUE', '', .) %>% 
    gsub(geno, '', .) %>% 
    gsub(year, '', .) %>% 
    gsub('TRUE', '', .)
  
  return(home_coef)
}


#' @title .two_tailed <internal>
#' 
#' @description perform two-tailed test on permutation values
#' 
#' @param x matrix. Each column is the result of a different permutation. The first column is the original data.
#' 
#' @return a vector of p-values for each row in x.

.two_tailed = function(x) {
  require(magrittr)
  
  alpha = sweep(x, 1, x[, 1], '>=') %>% 
    rowSums %>% 
    divide_by(ncol(x))
  p_val = cbind(alpha, 1-alpha) %>% 
    apply(1, min) %>% 
    multiply_by(2)
  return(p_val)
}

#' @title .calculate_intervals <internal>
#' 
#' @description calculate median and 90% confidence intervals of difference of permutation and observed data
#' 
#' @param x permutation results. Rows are values, columns are permutations. The first column is observed.
#' 
#' @return a matrix of columns "median", "q05" (5th percentile), "q95" (95th percentile) of (observed - permutation)
.calculate_intervals = function(x) {
  require(magrittr)
  
  difference = sweep(x, 1, x[, 1], '-') %>%
    multiply_by(-1)
  
  out = cbind(
    median = apply(difference, 1, median, na.rm=TRUE),
    p05 = apply(difference, 1, quantile, 0.05, na.rm=TRUE),
    p95 = apply(difference, 1, quantile, 0.95, na.rm=TRUE)
  )
  
  return(out)
}

#' @title permute_hfa
#' 
#' @description test the magnitude and significance of the home field advantage
#' versus what would be expected by chance.
#' 
#' @param data data.frame
#' @param level parameter at which to apply the home field advantage. Default is population.
#' @param site column name containing spatial environmental information
#' @param year column name containing temporal environmental information
#' @param geno column name containing genotype or variety information
#' @param pheno column name containing phenotype or performance information
#' @param popn column name differnetiating sub-populations of genotypes within the dataset.
#' 
#' @return 
#' 
permute_hfa = function(data,
                       level=c('population', 'genotype', 'year', 'site'),
                       site=NA, 
                       year=NA, 
                       geno=NA, 
                       pheno=NA, 
                       popn=NA,
                       times=99, 
                       blup_home=TRUE,
                       parallel=TRUE,
                       seed=NULL) {
  require(magrittr)
  require(parallel)  # will need to transition to foreach for compatibility
  
  ncpu = ifelse(parallel, detectCores(), 1)
  level = match.arg(level)
  LAPACK = !(level %in% c('site', 'year'))  # cannot guarantee a home site in each year, or for each variety.
  # LAPACK=TRUE
  
  # new column names
  part_pheno = paste0('part_', pheno)
  rel_pheno = paste0('rel_', pheno)
  
  # formula for calculating HFA effect
  ff = switch(level,
              'population' = ' ~ geno + is_home',  # overall HFA
              'genotype' = ' ~ geno + geno:is_home',
              'year' = '~ geno + year:is_home',
              'site' = '~ geno + site:is_home')  # HFA for each genotype
  ff %<>%
    gsub('geno', geno, .) %>% 
    gsub('year', year, .) %>% 
    gsub('site', site, .) %>% 
    formula
  
  # select and format data into list of dataframes for each populations
  dd = 
    c(site, year, geno, pheno, popn) %>% 
    na.omit %>% 
    data[, .]
  
  if (! (is.na(popn))) {
    dd %<>% split(dd[, popn])
  } else {
    dd %<>% list
  }
  
  
  id_home_site = colnames(data) %>% 
    grepl('is_home', .) %>% 
    any %>% 
    not
  if (id_home_site) {
    dd %<>% mclapply(function(x) {
      id_home(x, site, year, geno, pheno, blup=blup_home, verbose=FALSE)
    }, 
    mc.cores=ncpu)
  }
  
  # calculate partial and relative yields
  dd %<>% mclapply(function(x) {
    
    # calc rel_pheno and is_home
    x %<>% id_home(site, year, geno, pheno, blup_home, verbose=FALSE)

    # Partial phenotype based on site-year. Relatively slow. 
    x[, part_pheno] = 
      paste(pheno, '~', year, '*', site) %>% 
      formula %>% 
      .quick_resid(data=x) %>% 
      scale(scale=FALSE)
    
    return(x)
  }, mc.cores=ncpu)
  
  # permute HFA within each population
  results = lapply(dd, function(x) {
    # Set up structured permutations
    sets = .generate_sets(x, site, year, times, seed)
    
    # Permute HFA
    coef_permute = mclapply(sets, function(ss) {
      # ID home site
      x[, c(rel_pheno, part_pheno)] %<>% .[ss, ]  # permute phenotypes within site-year
      x %<>%
        split(x[, geno]) %>%
        lapply(.id_top_pheno, site, rel_pheno, blup=blup_home, verbose=FALSE) %>%  # This is the rate-limiting step.
        do.call(rbind, .)
      
      # calculate HFA using qr decomposition and lapack
      home_coef = .calculate_hfa(x, ff, part_pheno, geno, year, LAPACK)
      
      return(home_coef)
    }, mc.cores=ncpu) %>% 
      do.call(cbind, .)
    
    colnames(coef_permute) = c('observed', 
                               paste0('perm', 1:(ncol(coef_permute)-1)))
    
    # Calculate p-values and effects
    test = cbind(
      intervals = .calculate_intervals(coef_permute),
      p_val = .two_tailed(coef_permute)
    )
    
    results = list(results = test,
                   perms = coef_permute)
    
    return(results)
  })
  
  if (level == 'population') {
    test_results = lapply(results, function(x) x$results) %>% 
      do.call(rbind, .) %>% 
      set_rownames(names(results))
    perms = lapply(results, function(x) x$perms) %>% 
      do.call(rbind, .) %>% 
      set_rownames(names(results))
  } else {
    lvl = switch(level, 
                 'genotype' = geno,
                 'year' = year,
                 'site' = site)
    test_results = names(results) %>% 
      lapply(function(x) {
      res = results[[x]]$results
      res = data.frame(popn = x,
                       lvl = rownames(res),
                       res,
                       stringsAsFactors=FALSE) %>% 
        set_rownames(NULL)
      names(res) %<>% 
        gsub('lvl', lvl, .) %>% 
        gsub('popn', popn, .)
      return(res)
    }) %>% 
      do.call(rbind, .)
    
    perms = lapply(results, function(x) x$perms)
  }
  
  out = list(home_field = test_results,
             perms = perms)

  return(out)
}


```

```{r temporal_hfa}
#' @title calculate_temporal_hfa <internal>
#' 
#' @description calculate hfa for each year
#' 
#' @param x data.frame
#' @param ff formula
#' 
#' @details the formula should contain a term, <year>_num:is_home, and also specify a response (e.g. pheno)
#' 
#' @return a data.frame with the coefficients plus relevant 

.calculate_temporal_hfa = function(x, ff, year) {
  y = as.character(ff)[2]
  
  # get the coefficients
  home_coef = lm(ff, x) %>% 
    summary %>% 
    coef
  
  # format
  home_coef %<>% 
    rownames %>% 
    grepl(':is_homeTRUE', .) %>% 
    home_coef[., ]
  
  rownames(home_coef) %<>% 
    gsub(':is_homeTRUE', '', .) %>% 
    gsub(year, '', .)

  # return values
  home_coef %<>% 
    data.frame(year = rownames(.),
               year_num = type.convert(rownames(.)),
               .,
               p.adj = p.adjust(home_coef[, 'Pr(>|t|)']),
               stringsAsFactors=FALSE) %>% 
    set_rownames(NULL)
  colnames(home_coef) %<>% 
    gsub('year', year, .) %>% 
    gsub('Pr...t..', 'p.value', .) %>% 
    gsub('Std..Error', 'Std.Error', .)
  
  
  return(home_coef)
}

#' @title temporal_hfa
#' 
#' @description calculate trend in home field advantage across time. The formula is:
#' 
#' `pheno ~ year + site + geno + year:site + year:is_home`. 
#' 
#' The year:is_home coefficient is the home field advantage in each year. 
#' 
#' @param data data.frame
#' @param site column in `data` indicating spatial environment
#' @param year column in `data`indicating temporal environment
#' @param geno column in `data` indicating genotype
#' @param pheno column in `data` indicating phenotype
#' @param popn column in `data` indicating groups of genotypes
#' @param blup_home identify home site using blup?

temporal_hfa = function(data, site, year, geno, pheno, popn=NA, 
                        blup_home=TRUE, 
                        formula_override=NA,
                        parallel=TRUE) {
  require(magrittr)
  require(parallel)
  require(car)
  
  if (grepl('mingw', version$os)) parallel=FALSE
  ncpu = ifelse(parallel, detectCores(), 1)
  
  # set up dataframe
  
  if (is.na(popn)) {
    dd = list(data)
  } else {
    dd = split(data, data[, popn])
  }
  
  # formula
  if (is.na(formula_override)) {
    ff = "pheno ~ year + site + geno + year:site + year:is_home"  %<>%
      gsub('pheno', pheno, .) %>% 
      gsub('site', site, .) %>% 
      gsub('year', year, .) %>% 
      gsub('geno', geno, .) %>% 
      formula
  } else {
    ff = formula_override
  }
  
  # has home already been identified?
  id_home_site = colnames(data) %>% 
    grepl('is_home', .) %>% 
    any %>% 
    not
  if (id_home_site) {
    dd %<>% mclapply(function(x) {
      id_home(x, site, year, geno, pheno, blup=blup_home, verbose=FALSE)
    }, 
    mc.cores=ncpu)
  }
  
  # calculate annual hfa
  out = mclapply(dd, .calculate_temporal_hfa, ff, year,
                 mc.cores=ncpu) %>% 
    do.call(rbind, .)
  
  # format output
  # population, if specified
  if (!is.na(popn)) {
    popn_id = rownames(out) %>% 
      strsplit('\\.') %>% 
      sapply('[', 1)
    
    out = data.frame(
      popn = popn_id,
      out,
      stringsAsFactors=FALSE
    ) %>% 
      set_rownames(NULL)
    colnames(out) %<>% gsub('popn', popn, .)
  }
  
  # year/temporal
  if (is.factor(data[, year])) {
    out[, year] %<>% factor(levels=levels(data[, year]))
  }
  
  # run anova
  year_num = paste0(year, '_num')
  ff_aov = paste('Estimate', year_num, sep='~')
  if (!is.na(popn)) {
    ff_aov %<>% paste('*', popn)
  }
  
  out_lm = ff_aov %>% 
    formula %>% 
    lm(data=out)
  out_aov = Anova(out_lm)
  
  out = list(anova = out_aov,
             model = out_lm,
             temporal_hfa = out,
             formulas = c(hfa = ff,
                        anova = ff_aov))
  
  return(out)
}

```






## Example stuff

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
